Exploiting Intrinsic Noise to Achieve Heterogeneous Control of Cell Differentiation Dynamics in the Presence of Time Delays, Model Uncertainties, and Extrinsic Variability

### 2023-07-12 paper.tex
edit first half of trajectory section.

We analyzed the control performance over a domain of target points to find if some regions are easier to control than others. Stability analysis of the auto-regulated system using a moderate light input can create up to four stable points and one unstable point in the state space.  These points give rise to basins which may attract probability more easily than other points. Control performance of the stochastic system was analyzed over a domain of different set points and were optimized using more observable (FAC) or less observable (PAC) control.

Controllers were optimized over a two dimensional domain of target points $[T_1,T_2]$ between between five and forty-five using the FAC input method or the PAC input method. Figure \ref{CtrlP} (A) shows the optimized control input of the system when the target is $T=[20,25]$ after optimizing the FAC. Figure \ref{CtrlP} (B) show the steady state probability distribution of the controlled system.  Figure \ref{CtrlP} (C and D) show the optimized controller and probability distributions using the PAC method. The unstable point in the rate equation matches the unstable point near $(20,20)$ in this domain.

FAC and PAC control performance over the domain of static set points Fig\ref{CtrlP}(E) and (F) showed regions of good control in the FAC between 10 and 30 but a smaller region in the PAC. The FAC control performance was found to be symmetric along the dashed black line. The cause was found to be that symmetric pairs had control inputs which were exact transposes of each other. The unstable point of the system is known to be near [20,20] which is also a point of poor control for the FAC system with respect to local control performance. Edge regions were difficult to control, likely due to the geometry of how control inputs move the probability.

@@

In \ref{Methods}, we developed a method to control the dynamics of a slowly changing system on a predefined path $r(x)$ by cycling through controllers at a fixed frequency. Thirty-two controllers were optimized along a pathway in order to target an in-sync reference point \ref{CR}, a phase lagged reference point \ref{CR}, or frequency separated reference point\ref{CR}(A). The synchronous signal is given by Fig \ref{CR}(B1) $r_1(x)=[20 sin(2 \pi x)+10, 20 sin(2 \pi x)+10]$, the phase lagged signal Fig \ref{CR}(C1) is given by $r_2(x)=[20 sin(2 \pi x)+10, 20 cos(2 \pi x)+10]$, and the frequency changed signal Fig \ref{CR}(D1) is given by $r_3(x)=[20 sin(2 \pi x)+10, 20 sin(6 \pi x)+10]$ using 32 steps for each controller.

FSP simulations of cyclically driven systems were slowly driven at a frequency of $1e-4$ seconds and their medians are shown in the Fig \ref{CR}(B2, C2, and D2)(red line, blue line). The time series probability distributions of $x_1$ and $x_2$ can be seen as red clouds and blue clouds respectively. Regions with purple clouds are the exact overlap of red and blue. Error distributions of the system were taken by shifting the probability distribution from the reference point to the origin Fig \ref{CR}(B2, C2, and D2). Root mean squared error of each simulation are shown in Fig \ref{CR}(B3, C3,D3).

Simulations show that the synchronous control performed the best with an average RSME of $6.9$ despite moving over the unstable point in the system. When the system is driven with a phase-lag the average RSME of the score increases to $8.2$ and when driven at a different frequency the average RMSE goes up to $8.6$.  
Suprisingly, the system is able to effectively control two signals to two different trajectories at different phase and frequency and using only a single control input despite the presence of noise Fig \ref{CR}(D3). Medians of the marginal distributions show good match with the desired reference signal in each control scheme Fig \ref{CR}(B4,C4, and D4), and time series trajectories of the RMSE show spikes in the control performance using any control strategy.

@@

Slowly responding systems driven at high frequency often exhibit poor control performance which improves as the driving frequency matches the time scale of the system.  The combination of both frequency driving and noise could be catastrophic for system control performance. Conversely, increasing granularity($\alpha$) may improve control performance but alters the system's time scale. Therefore these competing effects of system scale has not yet been analyzed.

To examine how driving frequency and system scale affects control performance, Sixty-four SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 16 cycles after reaching steady state. For each $(\alpha,f)$ pair the average RMSE of the steady state trajectories were calculated.

>>>>

%We next analyzed the control performance over a domain of target points to find if some states are easier to achieve than others. For a two-cell system, stability analysis of the auto-regulated system using a moderate light input can create up to four stable points and one unstable point in the state space.  These points give rise to basins which may attract probability more easily than other points.
We next optimized the FAC controller and calculated its performance for the two-cell system over a two dimensional domain of target points $[T_1,T_2]$ between between five and forty-five. For example, Fig \ref{CtrlP} (A) shows the optimized FAC control input of the system when the target is $\mathbf{T}=[20,25]$, and Fig.\ \ref{CtrlP} (B) shows the corresponding steady state probability distribution of the controlled system.  Figure \ref{CtrlP} (C and D) show the the same thing but for a different target state of $\mathbf{T}=[XXX,XXX]$.

\brian[Let's remove the PAC from this section.  Show only the FAC and and replace the Fig 6C/D with the FAC at a different target point. Put a new star of a different color to match to the new 6C/D.  Remove Fig 6F.  Make sure to change all the figure labels and titles and captions.]

Figure \ref{CtrlP}(E) shows the FAC control performance over the entire domain of static set points, and illustrates that some regions are easier to obtain than others. For example, near the two stable fixed points at [XXX,XXX] and [XXX,XXX], the control performance is quite strong (white regions), whereas near the unstable point [20,20], the FAC results in relatively worse control.

Edge regions were difficult to control, likely due to the geometry of how control inputs move the probability. \brian[I don't understand this statement.]

@@

We developed a method (Section \ref{sec:Dynamic}) to control the system dynamics to follow a predefined path $r(x)$ by cycling through controllers optimized for individual states along that path. Specifically, we found 32 controllers along each three representative pathways, including an in-sync reference point (Fig.\ \ref{CR}B1), a phase lagged reference point (Fig.\ \ref{CR}C1),  and a  frequency separated reference point (Fig.\ \ref{CR}D1). The synchronous signal is given by Fig \ref{CR}(B1) $r_1(x)=[20 sin(2 \pi x)+10, 20 sin(2 \pi x)+10]$, the phase lagged signal Fig \ref{CR}(C1) is given by $r_2(x)=[20 sin(2 \pi x)+10, 20 cos(2 \pi x)+10]$, and the frequency changed signal Fig \ref{CR}(D1) is given by $r_3(x)=[20 sin(2 \pi x)+10, 20 sin(6 \pi x)+10]$ using 32 steps for each controller.

FSP simulations for these cyclically driven systems were slowly driven at a frequency of $1e-4$ seconds \brian[Check units. This doesn't make sense.  did you mean period?  If so, this is very fast.  Or did you mean s$^{-1}$?] and Fig.\ \ref{CR}(B2, C2, and D2) show their response distributions (shading) and median responses (lines) for the two cells x1 and x2 in red and blue, respectively. Regions with purple shading depict the exact overlap of red and blue when both cells have the same distribution of response. We observe that the single-input control system is able to effectively control the system to follow all three input trajectories, including when the two reference trajectories have different phases and frequencies (Fig \ref{CR}D3). 
For all reference trajectories, the medians of the marginal distributions show good match with the desired reference signal (Fig \ref{CR}B4,C4, and D4). To quantify the overall performance, Figs.\ \ref{CR}(B3, C3,D3) show the root mean squared error as a function of time,
and Figs.\ \ref{CR}(B2, C2, and D2) show the time-averaged error distribution for the system response relative to the time varying target. From these figures, we observe that all trajectories result in short RMSE spikes at times, and these correspond to short, transient periods when the controller passes through the regions of poor control depicted in Fig.\ \ref{CtrlP}(E). Overall, the simulations show that the synchronous control performed best with an average RSME of $6.9$ despite moving over the unstable point in the system. When the system is driven with a phase-lag, the average RSME of the score increases to $8.2$, and when driven at a different frequency, the average RMSE goes up to $8.6$.

@@

%Slowly responding systems driven at high frequency often exhibit poor control performance which improves as the driving frequency matches the time scale of the system.  The combination of both frequency driving and noise could be catastrophic for system control performance. 
A given control system generally cannot effectively respond to reference signals that change faster that the system's natural time scale. Since we saw previously that increasing granularity ($\alpha$) improves steady state control performance (Fig.\ \ref{Time}G) but also lengthens the systems time scale (Fig.\ \ref{Time}H), one should expect that granularity would also affect the types and speeds of signals to which the system can respond.
To examine how driving frequency and system scale affects control performance, 64 SSA trajectories of the phase-separated dynamic controller were simulated over a two dimensional domain of points $(\alpha,f)$ for 16 cycles after reaching steady state. For each $(\alpha,f)$ pair, the average RMSE of the steady state trajectories were calculated.

%Brian got here
